Master Thesis

### 2013-06-17 constraints.tex
Last night shift

\chapter{The Last Part}

In this chapter, we will attempt to incorporate accounts of several
linguistic phenomena in a single framework, highlight the modularity
issues this brings up and propose some solutions to them.

@@

We will start with an account of the paired grammatical words mechanism
for negation in French. Recall the EPTDs we have seen
in~\ref{ssec:frigram}, repeated here on Figure~\ref{fig:ig-neg-rep}.

\begin{figure}
  \centering
  \includegraphics[scale=0.25]{ig-neg.pdf}
  \caption{\label{fig:ig-neg-rep} Frigram EPTDs for a negative
    determiner and the paired grammatical word $ne$.}
\end{figure}

At a basic level, $aucun$ functions like any other determiner and $ne$
as any verb modifier. However, we want to encode the constraint that
whenever a verb is modified by $ne$, the modified verb will demand that
one of its arguments contains a negative determiner. Since we need to
make a distinction between phrases that contain negative determiners and
those that do not, we will need to provide different types for both (two
terms with the same type are indistinguishable w.r.t. syntactic
composition in a type-logical grammar such as ACGs). Our type for
$aucun$ will thus end up being $N\_NEG{=}F \limp ((NP\_NEG{=}T \limp S)
\limp S)$ instead of the usual $N \limp ((NP \limp S) \limp
S)$.\footnote{Strictly following the EPTD for $aucun$ would lead us to
  give it a type $DET\_NEG{=}F$ and then give the two types
  $DET\_NEG{=}F \limp ((NP\_NEG{=}F \limp S) \limp S)$ and $DET\_NEG{=}T
  \limp ((NP\_NEG{=}T \limp S) \limp S)$ to nouns. In this
  demonstration, we prefer to align our examples with the treatment
  which is customary to ACG research.} The type for $le$ will need to be
able handle cases where its argument noun either already contains a
negative determiner or not.

@@

The type of $ne$ will be more verbose, as it will consume a verb and
transform its valency so that it demands that exactly one of its $NP$
arguments contains a negative determiner.

@@

Furthermore, we could add the case when both the subject and the object
contain negative determiners, such as in~(\ref{ex:aucun2}). This case is
currently not covered by Frigram (though it could be), but is considered
grammatical by French speakers. Finally, in a complete grammar, $ne$
would have to provide types capable of transforming all the other
syntactic valencies.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.25]{parse-aucun.pdf}
  \caption{\label{fig:parse-aucun} One of the parse trees assigned to
    sentence (\ref{ex:aucun-obj}) by Frigram/Leopar.}
\end{figure}

However, this is not the only change we have to effect on our grammar in
order to properly handle the paired grammatical words for
negation. Consider the example sentence (\ref{ex:aucun-obj}) and the
parse tree assigned to it by Frigram on Figure~\ref{fig:parse-aucun}. It
is not enough that the embedded noun phrase $aucun\ tatou$ has a type
that tells us it contains a negative determiner. We need this bit of
internal information also at the level of the enclosing noun phrase
$l'odeur\ d'aucun\ tatou$ in which the former noun phrase is embedded
since this is a property of the enclosing noun phrase that is crucial to
its syntactic combinatorics. This presupposes the existence of a
mechanism for propagating this kind of information up the parse tree. In
the case of our ACG, this means that all the functions which modify
phrases must be aware whether or not their argument contains a negative
determiner and to convey this information in its result type as
well. Moreover, since the negative determiner can also come from a
complement of a prepositional phrase, the types of prepositions will
need to take into account the presence of negative determiners in both
their complements and the phrases they modify.

@@

@@

In sentence (\ref{ex:bad-neg-noneg}), we have no constant
$N_{ne_{tv_?}}$ for $ne$ which would yield a verb capable of consuming
two noun phrases that contain no negative determiners.

@@

There are two things that we would like to highlight about this
grammatical treatment. First off, we had to refine our types to make
them convey more than one piece of information in a single type
(e.g. something is a noun phrase and at the same time it is something
that contains a negative determiner). Second, a property that we wish to
express in the type of a term may be due to one of its subconstituents
and it is therefore necessary to encode the propagation of this property
up the parse tree in the types of all the intervening operators. In our
case, it meant that handling negative determiners and the paired
grammatical word for negation required us to not only define and type
constants for these two categories, but to also change the types of noun
modifiers ($que\ ...$) and prepositions ($de$).

@@

We will now turn to another phenomenon, extraction in relative clauses,
focus on one of its properties, the fact that subjects can be extracted
using $qui$ only from the relative clause itself whereas objects can be
extracted using $que$ from embedded clauses contained therein, and
present an implementation of this constraint by way of types in an
ACG. The solution that we will show was published in
\cite{pogodalla2012controlling}. Our presentation is slightly adapted to
conform to the patterns seen in previous examples and it does not use
dependent types as they are not essential to this specific constraint.

@@

Now all of the infrastructure is in place for us to express the type of
the relative pronouns $qui$ and $que$.

@@

The above types ensure that the abstracted variables corresponding to
traces of extracted constituents will be given the type $NP\_VAR{=}T$,
that $qui$ can only be used to extract constituents from root positions
while $que$ allows for extraction from both root and embedded
positions.\footnote{NB: The constraint that $qui$ can only extract
  subjects and that $que$ can only extract objects is not enforced
  here. Likewise, this fragment does not handle multiple extraction.}

@@

While we have introduced the notion of an $NP\_VAR{=}T$ for describing
traces, we can cover another element of our running fragment, the
preposition $de$, and enforce the constraint that $qui$/$que$ cannot
extract prepositional complements by providing only a single type for
$de$ that accepts only unmoved noun phrases as complements.

@@

Given this type system, we can assign the type $S\_EXT{=}NO$ to the term
expressing the syntactic structure of (\ref{ex:good-ext}) but we cannot
a find a typable term for sentence (\ref{ex:bad-ext}), where $qui$ is
used to extract a subject from a clause embedded in a relative clause.
The problem with sentence (\ref{ex:bad-ext}) is that the type of the
relative clause is $NP\_VAR{=}T \limp S\_EXT{=}EMB$ which is not a valid
argument for any constant representing the relative pronoun
$qui$. Typing the variable $t$ with $NP\_VAR{=}F$ instead of
$NP\_VAR{=}T$ and using $E_{aime_1}$ instead of $E_{aime_2}$ would not
help, since the resulting relative clause would then have type
$NP\_VAR{=}F \limp S\_EXT{=}NO$, which would not be admitted by any
relative pronoun.

@@

\begin{align*}
A_{Marie} &: NP\_NUM{=}SG \\
A_{tatou} &: N\_NUM{=}SG \\
A_{tatous} &: N\_NUM{=}PL \\
A_{le} &: N\_NUM{=}SG \limp ((NP\_NUM{=}SG \limp S) \limp S) \\
A_{les} &: N\_NUM{=}PL \limp ((NP\_NUM{=}PL \limp S) \limp S) \\
A_{dort} &: NP\_NUM{=}SG \limp S \\
A_{dorment} &: NP\_NUM{=}PL \limp S \\
A_{aime_1} &: NP\_NUM{=}SG \limp NP\_NUM{=}SG \limp S \\
A_{aime_2} &: NP\_NUM{=}SG \limp NP\_NUM{=}PL \limp S \\
A_{aiment_1} &: NP\_NUM{=}PL \limp NP\_NUM{=}SG \limp S \\
A_{aiment_2} &: NP\_NUM{=}PL \limp NP\_NUM{=}PL \limp S \\
A_{qui_1} &: (NP\_NUM{=}SG \limp S) \limp N\_NUM{=}SG \limp N\_NUM{=}SG \\
A_{qui_2} &: (NP\_NUM{=}PL \limp S) \limp N\_NUM{=}PL \limp N\_NUM{=}PL \\
A_{de_1} &: NP\_NUM{=}SG \limp N\_NUM{=}SG \limp N\_NUM{=}SG \\
A_{de_2} &: NP\_NUM{=}PL \limp N\_NUM{=}SG \limp N\_NUM{=}SG \\
A_{de_3} &: NP\_NUM{=}SG \limp N\_NUM{=}PL \limp N\_NUM{=}PL \\
A_{de_4} &: NP\_NUM{=}PL \limp N\_Pl \limp N\_NUM{=}PL
\end{align*}

Given the above signature, we can assign the type $S$ to the syntactic
term of sentence (\ref{ex:good-agr}) but not to the one of sentence
(\ref{ex:bad-agr}). In the latter sentence, the type of the variable $t$
has to be $NP\_NUM{=}SG$ since it is used as an argument to $A_{dort}$. This
means that the relative clause $dort$ has type $NP\_NUM{=}SG \limp
S$. However, there is no type for $qui$ which would allow us to modify
the plural noun $A_{tatous}$ of type $N\_NUM{=}PL$ by a relative clause with a
missing singular noun phrase, which has the type $NP\_NUM{=}SG \limp S$.

@@

The individual types will have to respect all of the constraints at the
same time, each one being a complete specification of a possible
situation (complete w.r.t. the features listed above). To see and
appreciate what this means in practice, we give the types for the
preposition $de$:

@@

This highlights two important problems with the current approach. One is
the untenable complexity. Even though the three mechanisms are
completely independent of each other, here we are forced to refer to all
of them in every type. It is no longer easy to glance at the types
assigned to $de$ and see how it behaves with respect to, for example,
the mechanism of negation (compare with the type assignments we gave to
$de$ in previous subsections). Another cause of unreadability of this
signature is due to the sheer number of types on display, which leads us
to our second issue.

@@

While extending the type system by including dependent types is
definitely useful and it helps to solve the problem of having our
signatures grow exponentially in size, it does not solve our problem
with the complexity of the new grammar. Granted, the new meta-type
written in the dependent style is more readable and easier to reason
about than the enumeration of 12 basic types we presented at the
beginning of this section, however, the result is still complex in
that it forces us to describe the behavior of three independent
mechanisms in one place. This one place happens to be a wordform which
has seemingly little connection to these mechanisms, the preposition
$de$. We can therefore imagine that it will participate in many other
phenomena and that the type we assign to it will grow beyond our
abilities to reason about it effectively.

Furthermore, the current notation does not even make it obvious that
these are independent mechanisms. One has to examine the type to make
sure that there is no interaction. In the rest of this chapter, we will
devote ourselves to investigating ways of disentangling these.

@@

\section{Intersections of Languages}
\label{sec:sects-of-langs}

We will turn to another solution to the problem of complex grammars that
try to implement multiple independent constraints. This solution will be
based on a generalization of ACGs, but in order to introduce this
generalization, we will first need to present the manner in which ACGs
and their compositions are formulated in ACG literature.

\subsection{The Usual Formalization of ACGs}
\label{ssec:usual-acgs}

Let us consider a system with three signatures $\Sigma_1$, $\Sigma_2$
and $\Sigma_3$, their respective distinguished types $S_1$, $S_2$ and
$S_3$ and two lexicons $\mathcal{L}_{12} : \Sigma_1 \to \Sigma_2$ and
$\mathcal{L}_{23} : \Sigma_2 \to \Sigma_3$ such that
$\mathcal{L}_{12}(S_1) = S_2$ and $\mathcal{L}_{23}(S_2) = S_3$. See
Figure~\ref{fig:acg-serial-comp} for a diagrammatic visualization of
this structure.

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{diagrams/acg-serial-comp.pdf}
    \caption{\label{fig:acg-serial-comp} Serial composition of ACGs.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{diagrams/acg-parallel-comp.pdf}
    \caption{\label{fig:acg-parallel-comp} Parallel composition of ACGs.}
  \end{subfigure}
  \caption{\label{fig:acg-comp-modes} Two modes of composition for
    ACGs.}
\end{figure}

We can now define languages $A_1$, $A_2$ and $A_3$ where $A_i$ is the
set of terms from $\Lambda(\Sigma_i)$ that have the type $S_i$. Such
languages, which are defined only in terms of a signature and some
distinguished type, are called \emph{abstract languages} in the
terminology of ACGs. They consist of terms whose structure is
constrained only by the types assigned by the signature.

Further than that, we can look at the languages $\mathcal{L}_{12}(A_1)$
and $\mathcal{L}_{23}(A_2)$, images of the abstract languages given by
the lexicons. Since the lexicons are homomorphic mappings w.r.t. the
types, all elements of $\mathcal{L}_{12}(A_1)$ will have the type
$\mathcal{L}_{12}(S_1) = S_2$ and will therefore also belong to $A_2$
(similarly, we have that $\mathcal{L}_{23}(A_2) \subseteq A_3$). These
new languages, which are formed by mapping an ACG abstract language
using a lexicon, are called \emph{object languages}. Their terms are
constrained not only by the types of the signature they are built upon
but also by the type system of the abstract signature.

Finally, we can also take the composition of lexicons $\mathcal{L}_{13}
= \mathcal{L}_{23} \circ \mathcal{L}_{12}$, which itself is also a
lexicon, and map it over $A_1$. This gives us a language
$\mathcal{L}_{13}(A_1) = \mathcal{L}_{23}(\mathcal{L}_{12}(A_1))$ which
is at the same time the image of the object language
$\mathcal{L}_{12}(A_1)$ given by $\mathcal{L}_{23}$ and also the image
of the abstract language $A_1$ given by $\mathcal{L}_{13}$ (thus it is
itself also an object language). This points to an interesting property
of ACG languages: they are closed under transformation by a lexicon.

In the ACG literature, ACGs are defined as quadruples $\mathcal{G} =
\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$ where
$\Sigma_A$ is the abstract signature, $\Sigma_O$ is the object
signature, $\mathcal{L} : \Sigma_A \to \Sigma_O$ is a lexicon linking
the two and $S$ is a distinguished type of the abstract signature
$\Sigma_A$. From this grammar, we can define two languages. The abstract
language $\mathcal{A}(\mathcal{G}) = \{t \in \Lambda(\Sigma_A) \mid
\ \vdash_{\Sigma_A} t : S\}$ and the object language
$\mathcal{O}(\mathcal{G}) = \mathcal{L}(\mathcal{A}(\mathcal{G})) = \{t
\in \Lambda(\Sigma_O) \mid \exists u \in \mathcal{A}(\mathcal{G}),
\mathcal{L}(u) = t\}$. The composition of two ACGs $\mathcal{G}_1 =
\mathopen{<}\Sigma_1, \Sigma_2, \mathcal{L}_{12}, S_1\mathclose{>}$ and
$\mathcal{G}_2 = \mathopen{<}\Sigma_2, \Sigma_3, \mathcal{L}_{23},
S_2\mathclose{>}$ is defined as $\mathcal{G}_2 \circ \mathcal{G}_1 =
\mathopen{<}\Sigma_1, \Sigma_3, \mathcal{L}_{23} \circ \mathcal{L}_{12},
S_1\mathclose{>}$.

The definition of ACGs given above is quite practical in that it is
complete, i.e. it gives a precise account of what set of languages are
generated by ACGs (abstract and object languages for which there exist
the quadruples described above). This contrasts with the presentation of
ACGs that we gave in~\ref{sec:acg}, where we defined ACGs as collections
of signatures associated with distinguished types and connected with
lexicons. Our definition only introduced the requisite machinery, but
delayed the delimitation of the way languages can be defined. Leaving
this question open will let us answer it now by providing an extension
of ACGs which will help us overcome the modularity issues that have been
the theme of this chapter.\footnote{Incidentally, this was not the
  reason for the way we presented ACGs in~\ref{sec:acg}. We chose this
  style as it grouped together the information in a more natural way. We
  approach signatures as descriptions of domains (strings, syntax,
  semantics), each one being associated with a distinguished type
  telling us which objects the domain has set out to describe (strings,
  sentences, truth values). Having distinguished types directly attached
  to signatures allows us to omit them from the ACG quadruples,
  $\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L},
  \cancel{S}\mathclose{>}$. However, this only leaves us with triples
  containing the two signatures and the lexicon connecting them, and
  since we can infer the signatures themselves from the domain and the
  range of the lexicon $\mathcal{L} : \Sigma_A \to \Sigma_O$, we do not
  have much motivation to introduce these tuples,
  $\mathopen{<}\cancel{\Sigma_A}, \cancel{\Sigma_O}, \mathcal{L},
  \cancel{S}\mathclose{>}$. Instead of composing ACGs, we compose just
  the lexicons.

On top of that, we will want to define signatures and lexicons to yield
multiple different object languages (at least one for the string
representations of forms and one for the semantic representations of
meaning), which would mean having at least two different ACGs which are
somehow linked together (e.g. by sharing the same abstract
signature). However, since these two grammars are so interrelated, we
would like to reason about them in a way that makes these relationships
explicit and thus easier to study as opposed to considering a collection
of grammars which are somehow implicitly similar.}

\subsection{Patterns of Composition in ACGs}
\label{ssec:acg-patterns}

Before we turn to our proposal, we will quickly recall some of the other
styles of composition used in ACGs (a more detailed exposition can be
found in \cite{pogodalla2012controlling}). Besides making the object
signature of one grammar be the abstract signature of another as we saw
before (Figure~\ref{fig:acg-serial-comp}), we can have two grammars
share the same abstract signature allowing us to transduce between terms
in the two object signatures (Figure~\ref{fig:acg-parallel-comp}).

NB: \emph{Transduction} is a process in which mapping an object $o_1$ to
(potentially many) other objects $o_2$ is realized by first inverting
one function $f_1$ to find the object's antecedents $a$ and then
applying to them another function $f_2$, i.e. an object $o_1$ is
transduced to an object $o_2$ whenever there exists an $a$ such that
$f_1(a) = o_1$ and $f_2(a) = o_2$. In our example, terms from
$\mathcal{O}( \mathopen{<} \Sigma_1, \Sigma_2, \mathcal{L}_{12}, S
\mathclose{>} )$ are transduced to terms from $\mathcal{O}( \mathopen{<}
\Sigma_1, \Sigma_3, \mathcal{L}_{13}, S \mathclose{>} )$ by inverting
the lexicon $\mathcal{L}_{12}$ to find an antecedent in $\mathcal{A}(
\mathopen{<} \Sigma_1, \Sigma_2, \mathcal{L}_{12}, S \mathclose{>} )$
and then applying to it the lexicon $\mathcal{L}_{13}$.

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{diagrams/parallel-over-serial.pdf}
    \caption{\label{fig:parallel-over-serial} Transduction from an
      abstract signature.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{diagrams/serial-over-parallel.pdf}
    \caption{\label{fig:serial-over-parallel} Constraint on an abstract
      signature.}
  \end{subfigure}
  \caption{\label{fig:acg-comp-patterns} Mixing the ACG modes of
    composition.}
\end{figure}

These two modes of composition can be fruitfully mixed. Consider the
graph on Figure~\ref{fig:parallel-over-serial} in which we introduce a
syntactic signature $Syntax$ where different terms are only allowed to
yield the same string whenever they represent different parses of a
syntactically ambiguous phrase. This means that purely semantic
distinctions such as scope ambiguity have to be moved to a more abstract
signature ($SyntSem$) from whose terms the semantic representation can
be derived by a function ($\mathcal{L}_{InterSem}$). This approach is
studied more deeply in \cite{pogodalla2007generalizing}.

Another interesting composition pattern is to take an existing structure
of grammars and add a new abstract signature on top
(Figure~\ref{fig:serial-over-parallel}). This new signature ($Constr$)
can be used to further constrain the items in the previously
abstract-most signature ($Synt$). In \cite{pogodalla2012controlling},
this is what the authors use to develop their constraints on
extraction. Each of their constraints is presented as a new abstract
signature constraining the original syntactic signature.

However, a question that the authors do not pose, but which is of great
interest to us, is how to combine all these constraints. If we were to
apply the same pattern repeatedly to get a structure like the one on
Figure~\ref{fig:stacked-constraints}, we would have to deal with
successively more complex type systems on each level. This is due to the
condition that a lexicon has to be a homomorphism and thus it must,
among other things, always map well-typed terms to well-typed
terms. Because of this, the terms in the abstract language of the
signature we introduce to handle the second constraint must also satisfy
the first constraint, since mapping them by $\mathcal{L}_{Constr_2}$
must yield a well-typed term in the signature corresponding to the first
constraint. In the end, whenever we wish to add a new signature to
handle another constraint, we have to acknowledge and reimplement in it
all of the existing constraints. This effectively renders the pattern
useless since we might just as well consider the final abstract-most
signature handling all the constraints ($Constr_{1 \land 2}$) and attach
it directly to the syntactic signature ($Synt$) by taking the
composition of the lexicons in between ($\mathcal{L}_{Constr_1} \circ
\mathcal{L}_{Constr_2}$).

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{diagrams/stacked-constraints.pdf}
    \caption{\label{fig:stacked-constraints} Stacking the constraints
      vertically.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{diagrams/constraints-side-by-side.pdf}
    \caption{\label{fig:constraints-side-by-side} Laying out the
      constraints side by side.}
  \end{subfigure}
  \caption{Constraining a syntactic signature with multiple
    constraints.}
\end{figure}

What we would like instead is to put the signatures side by side such
that they do not depend on each other
(Figure~\ref{fig:constraints-side-by-side}). However, our language of
syntactic structures is no longer defined by mapping a single abstract
language through a lexicon. It is outside of the scope of the common
formalization of ACGs that we presented in~\ref{ssec:usual-acgs}. How do
we define this language in a manner that is consistent with our current
definitions of ACGs? We can arrive at the answer by examining how the
graphical representations of ACGs are connected to the languages they
define.

\subsection{Interpreting ACG Diagrams}

We will associate a language to every node of an ACG diagram. Before we
begin, we notice that every ACG diagram formed by the above modes of
composition is an arborescence\footnote{An arborescence is a directed
  rooted tree, where every edge points away from the root.} and
therefore every node has at most one predecessor (the parent). To the
root of the tree, the abstract-most signature in the graph, we will
assign the abstract language generated by that signature and its
distinguished type. To all the other nodes, we will assign the result of
mapping their parent's language by the lexicon linking the two. By
homomorphism of the lexicons, we can find distinguished types for all
the non-root nodes. They are simply the images of their parents'
distinguished types by the connecting lexicons. For all of the nodes, it
is true that every term in the language assigned to the node has the
node's distinguished type.

We can confirm that the above graphical model of defining ACGs is
coherent.

First, we can show that our interpretation is sound meaning that any
arborescence adorned with signatures on the nodes, compatible lexicons
on the edges and a distinguished type on the root can only define ACG
languages (abstract or object languages). This is trivial, since the
language corresponding to the root is an ACG abstract language and the
class of ACG languages is closed under transformation by a lexicon and
therefore all of the languages corresponding to the descendants of the
root are ACG languages as well.

Second, we can try and show some kind of completeness. However, there is
not much formalization done to describe a ``system of ACGs that share
some common signatures'' (something that our generalization will
fix). We can nevertheless show that it is possible to express any ACG
using our graphical model. Given a grammar $\mathcal{G} =
\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$, we just
produce a tree whose root has the signature $\Sigma_A$ and distinguished
type $S$ and which is connected to a single child node having the
signature $\Sigma_O$ by an edge labelled by the lexicon
$\mathcal{L}$. Then the two nodes of our tree define both the abstract
and the object language defined by $\mathcal{G}$. Similarly, we could
argue that the languages we aim to define using ACG diagrams (such as
the ones on Figures \ref{fig:acg-comp-modes} and
\ref{fig:acg-comp-patterns}) are covered by our assignment of languages
to nodes of the diagrams.

We have now given a formal account of how we can read language
definitions from ACG diagrams. Still, we have kept our analysis to the
arborescent diagrams used in current ACG literature, where every node
has at most one parent. We will now proceed to generalize it to cases
where nodes have more than just one parent.

To make the generalization more visible, we will switch from an
algebraic manner of presentation to a relational one. Before, we stated
that the language of a node is the image of its parent's language.

$$
Language(u) = Lexicon(Language(Parent(u)))
$$

Now, we will restate this as saying that the language of a node is the
set of terms that have an antecedent in the parent's language.

$$
Language(u) = \{x \in \Lambda(Signature(u)) \mid \exists a \in
Language(Parent(u)), \; Lexicon(a) = x\}
$$

This definition works for non-root nodes of an ACG diagram. In the root
case, we have to constrain the language by stating that that the terms
of the language must have the root's distinguished type. This is a
property that is also true in the non-root cases due to the homomorphic
properties of lexicons. If we wanted to unify the definitions of the
root language and the non-root languages, we would have to check that a
term of the language has the distinguished type and that it has an
antecedent in the parent, if it has any parent. A relational statement
of this unification could look like this.

>>>>

\chapter{Treatment of Multiple Linguistic Constraints}
\label{chap:constraints}

In this chapter, we will attempt to incorporate accounts of several
linguistic phenomena in a single framework, highlight the modularity
issues this brings up.

@@

We will start with an account of the paired grammatical words mechanism
for negation in French, specifically the interaction between negative
noun phrases and the particle \emph{ne}. Recall the EPTDs we have seen
in~\ref{ssec:frigram}, repeated here on Figure~\ref{fig:ig-neg-rep}.

\begin{figure}
  \centering
  \includegraphics[scale=0.25]{images/ig-neg.pdf}
  \caption{\label{fig:ig-neg-rep} Frigram EPTDs for a negative
    determiner and the paired grammatical word \emph{ne}.}
\end{figure}

At a basic level, \emph{aucun} functions like any other determiner and
\emph{ne} as any verb modifier. However, we want to encode the
constraint that whenever a verb is modified by \emph{ne}, the modified
verb will demand that one of its arguments contains a negative
determiner. Since we need to make a distinction between phrases that
contain negative determiners and those that do not, we will need to
provide different types for both (two terms with the same type are
indistinguishable w.r.t. syntactic composition in a type-logical grammar
such as ACGs). Our type for \emph{aucun} will thus end up
being\footnote{Strictly following the EPTD for \emph{aucun} would lead
  us to give it a type $DET\_NEG{=}F$ and then give the two types
  $DET\_NEG{=}F \limp ((NP\_NEG{=}F \limp S) \limp S)$ and $DET\_NEG{=}T
  \limp ((NP\_NEG{=}T \limp S) \limp S)$ to nouns. In this
  demonstration, we prefer to align our examples with the treatment
  which is customary to ACG research.} $$N\_NEG{=}F \limp ((NP\_NEG{=}T
\limp S) \limp S)$$ instead of the usual $$N \limp ((NP \limp S) \limp
S)$$ The type for \emph{le} will need to be able handle cases where its
argument noun either already contains a negative determiner or not.

@@

The type of \emph{ne} will be more verbose, as it will consume a verb
and transform its valency so that it demands that exactly one of its
$NP$ arguments contains a negative determiner.

@@

Furthermore, we could add the case when both the subject and the object
contain negative determiners, such as in~(\ref{ex:aucun2}). This case is
currently not covered by Frigram (though it could be), but is considered
grammatical by French speakers. Finally, in a complete grammar,
\emph{ne} would have to provide types capable of transforming all the
other syntactic valencies.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.25]{images/parse-aucun.pdf}
  \caption{\label{fig:parse-aucun} One of the parse trees assigned to
    sentence (\ref{ex:aucun-obj}) by Frigram/Leopar.}
\end{figure}

However, this is not the only change we have to effect on our grammar in
order to properly handle the paired grammatical words for
negation. Consider the example sentence (\ref{ex:aucun-obj}) and the
parse tree assigned to it by Frigram on Figure~\ref{fig:parse-aucun}. It
is not enough that the embedded noun phrase \emph{aucun tatou} has a
type that tells us it contains a negative determiner. We need this bit
of internal information also at the level of the enclosing noun phrase
\emph{l'odeur d'aucun tatou} in which the former noun phrase is embedded
since this is a property of the enclosing noun phrase that is crucial to
its syntactic combinatorics. This presupposes the existence of a
mechanism for propagating this kind of information up the parse tree. In
the case of our ACG, this means that all the functions which modify
phrases must be aware whether or not their argument contains a negative
determiner and to convey this information in its result type as
well. Moreover, since the negative determiner can also come from a
complement of a prepositional phrase, the types of prepositions will
need to take into account the presence of negative determiners in both
their complements and the phrases they modify.

@@

In the above types, we not only propagate information about the presence
of a negative determiner, we also introduce a constraint in the
mechanism stating that it cannot be the case that there is a negative
determiner both in the modified noun and in its noun phrase complement
(i.e. phrases like \emph{aucun tatou d'aucun homme} are not acceptable).

@@

In sentence (\ref{ex:bad-neg-noneg}), we have no constant
$N_{ne_{tv_?}}$ for \emph{ne} which would yield a verb capable of
consuming two noun phrases that contain no negative determiners.

@@

There are two things that we would like to highlight about this
grammatical treatment. First off, we had to refine our types to make
them convey more than one piece of information in a single type
(e.g. something is a noun phrase and at the same time it is something
that contains a negative determiner). Second, a property that we wish to
express in the type of a term may be due to one of its subconstituents
and it is therefore necessary to encode the propagation of this property
up the parse tree in the types of all the intervening operators. In our
case, it meant that handling negative determiners and the paired
grammatical word for negation required us to not only define and type
constants for these two categories, but to also change the types of noun
modifiers (\emph{que ...}) and prepositions (\emph{de}).

@@

We will now turn to another phenomenon, extraction in relative clauses,
focus on one of its properties, the fact that subjects can be extracted
using \emph{qui} only from the relative clause itself whereas objects
can be extracted using \emph{que} from embedded clauses contained
therein, and present an implementation of this constraint by way of
types in an ACG. The solution that we will show was published in
\cite{pogodalla2012controlling}. Our presentation is slightly adapted to
conform to the patterns seen in previous examples and it does not use
dependent types as they are not essential to this specific constraint.

@@

Now all of the infrastructure is in place for us to express the type of
the relative pronouns \emph{qui} and \emph{que}.

@@

The above types ensure that the abstracted variables corresponding to
traces of extracted constituents will be given the type $NP\_VAR{=}T$,
that \emph{qui} can only be used to extract constituents from root
positions while \emph{que} allows for extraction from both root and
embedded positions.\footnote{NB: The constraint that \emph{qui} can only
  extract subjects and that \emph{que} can only extract objects is not
  enforced here. Likewise, this fragment does not handle multiple
  extraction.}

@@

While we have introduced the notion of an $NP\_VAR{=}T$ for describing
traces, we can cover another element of our running fragment, the
preposition \emph{de}, and enforce the constraint that
\emph{qui}/\emph{que} cannot extract prepositional complements by
providing only a single type for \emph{de} that accepts only unmoved
noun phrases as complements.

@@

Given this type system, we can assign the type $S\_EXT{=}NO$ to the term
expressing the syntactic structure of (\ref{ex:good-ext}) but we cannot
a find a typable term for sentence (\ref{ex:bad-ext}), where \emph{qui}
is used to extract a subject from a clause embedded in a relative
clause.  The problem with sentence (\ref{ex:bad-ext}) is that the type
of the relative clause is $NP\_VAR{=}T \limp S\_EXT{=}EMB$ which is not
a valid argument for any constant representing the relative pronoun
\emph{qui}. Typing the variable $t$ with $NP\_VAR{=}F$ instead of
$NP\_VAR{=}T$ and using $E_{aime_1}$ instead of $E_{aime_2}$ would not
help, since the resulting relative clause would then have type
$NP\_VAR{=}F \limp S\_EXT{=}NO$, which would not be admitted by any
relative pronoun.

@@

\begin{align*}
A_{Marie} &: NP\_NUM{=}SG \\
A_{tatou} &: N\_NUM{=}SG \\
A_{tatous} &: N\_NUM{=}PL \\
A_{le} &: N\_NUM{=}SG \limp ((NP\_NUM{=}SG \limp S) \limp S) \\
A_{les} &: N\_NUM{=}PL \limp ((NP\_NUM{=}PL \limp S) \limp S) \\
A_{dort} &: NP\_NUM{=}SG \limp S \\
A_{dorment} &: NP\_NUM{=}PL \limp S \\
A_{aime_1} &: NP\_NUM{=}SG \limp NP\_NUM{=}SG \limp S \\
A_{aime_2} &: NP\_NUM{=}SG \limp NP\_NUM{=}PL \limp S \\
\end{align*}
\begin{align*}
A_{aiment_1} &: NP\_NUM{=}PL \limp NP\_NUM{=}SG \limp S \\
A_{aiment_2} &: NP\_NUM{=}PL \limp NP\_NUM{=}PL \limp S \\
A_{qui_1} &: (NP\_NUM{=}SG \limp S) \limp N\_NUM{=}SG \limp N\_NUM{=}SG \\
A_{qui_2} &: (NP\_NUM{=}PL \limp S) \limp N\_NUM{=}PL \limp N\_NUM{=}PL \\
A_{de_1} &: NP\_NUM{=}SG \limp N\_NUM{=}SG \limp N\_NUM{=}SG \\
A_{de_2} &: NP\_NUM{=}PL \limp N\_NUM{=}SG \limp N\_NUM{=}SG \\
A_{de_3} &: NP\_NUM{=}SG \limp N\_NUM{=}PL \limp N\_NUM{=}PL \\
A_{de_4} &: NP\_NUM{=}PL \limp N\_NUM{=}PL \limp N\_NUM{=}PL
\end{align*}

Given the above signature, we can assign the type $S$ to the syntactic
term of sentence (\ref{ex:good-agr}) but not to the one of sentence
(\ref{ex:bad-agr}). In the latter sentence, the type of the variable $t$
has to be $NP\_NUM{=}SG$ since it is used as an argument to
$A_{dort}$. This means that the relative clause \emph{dort} has type
$NP\_NUM{=}SG \limp S$. However, there is no type for \emph{qui} which
would allow us to modify the plural noun $A_{tatous}$ of type
$N\_NUM{=}PL$ by a relative clause with a missing singular noun phrase,
which has the type $NP\_NUM{=}SG \limp S$.

@@

The individual types will have to respect all of the constraints at the
same time, each one being a complete specification of a possible
situation (complete w.r.t. the features listed above). To see and
appreciate what this means in practice, we give the types for the
preposition \emph{de}:

@@

This highlights two important problems with the current approach. One is
the untenable complexity. Even though the three mechanisms are
completely independent of each other, here we are forced to refer to all
of them in every type. It is no longer easy to glance at the types
assigned to \emph{de} and see how it behaves with respect to, for
example, the mechanism of negation (compare with the type assignments we
gave to \emph{de} in previous subsections). Another cause of
unreadability of this signature is due to the sheer number of types on
display, which leads us to our second issue.

@@

While extending the type system by including dependent types is
definitely useful and it helps to solve the problem of having our
signatures grow exponentially in size, it does not solve our problem
with the complexity of the new grammar. Granted, the new meta-type
written in the dependent style is more readable and easier to reason
about than the enumeration of 12 basic types we presented at the
beginning of this section, however, the result is still complex in that
it forces us to describe the behavior of three independent mechanisms in
one place. This one place happens to be a wordform which has seemingly
little connection to these mechanisms, the preposition \emph{de}. We can
therefore imagine that it will participate in many other phenomena and
that the type we assign to it will grow beyond our abilities to reason
about it effectively.

Furthermore, the current notation does not even make it obvious that
these are independent mechanisms. One has to examine the type to make
sure that there is no interaction.

TODO: Change this after splitting the chapters.

In the rest of this chapter, we will devote ourselves to investigating
ways of disentangling these.

@@
